
## Module 5: Exercises

In these exercises we are going to use a different survey dataset. It comes from the same series of surveys as the example you have been looking at so far, but it was taken one month earlier. There are many questions which are the same - `Gender`, `Age`, `AgeGrp`, `PoliticalParty`, `ApproveTrump`, `Education` and `Race`, although the respondents are different.
There are also two new columns for questions which only appeared in this particular survey

`Books`: "How many books have you read in the past year?"
`Ghosts`: "Do you believe in ghosts?"

If you have opened up this file without first unzipping all of the files into a single folder, and creating a Project file for this exercises, then please stop right now. Do those things!

Now you are back you can install any of the libraries used in this session that you might not already have. You should already have ggplot2 and dplyr; but you probably will now also need to install janitor.

Once you have done that, you are now ready to go and running the chunk below should load the libraries and import the data to be used in these exercises.

```{r setup}

library(ggplot2)
library(dplyr)
library(janitor)

SurveyData<-read.csv("SurveyData.csv")


```


## Exercise 1

**I would like to investigate the hypothesis that Democrats, on average, read more books in the past year than republicans**

*1.1 - Firstly - create a subset of the data which only includes Democrats and Republicans, and excludes the respondents who responded as Independent. Assign this to a new object called `DemRep`*
```{r}

```

*1.2 Now calculate the mean, median and standard deviation of the number of books read by political party*
```{r}


```

*1.3 Now produce histograms of the distribution of number of books read by political party*
```{r}

```

*1.4 Based on the summary statistics and plots, choose an appropriate statistical method to test the null hypothesis that there is no difference in the number of books read by political party.*
*Conduct the test and interpret the results*

```{r}

```



## Exercise 2

**I am now (for some reason or other) interested in finding out if there is a relationship between whether a respondent's gender and whether or not they believe in ghosts?**

*2.1 Firstly, produce a cross-tabulation of approval of gender and whether or not a respondent believes in ghosts.*
```{r}


```

*2.2 Convert this table into a sensible set of percentages (and format the numbers so they appear as percentages)*
```{r}


```



*2.3 Produce an appropriate graph displaying the relationship between gender and the belief in ghosts*
```{r}


```


*2.4 Based on the cross-tab select an appropriate statistical method to test the null hypothesis that there is no relationship between these variables. Conduct the test, and interpret the results*

```{r}

```


## Exercise 3

*Investigate the relationship between age and belief in ghosts. Produce some summary statistics and plots, and conduct a hypothesis test.*

Work through this whole process yourself. Make sure you add comments and explanations to yourself as you go through to explain decisions you make, and interpret the output you obtain.

